Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-08-30 paper/abstract.tex
some fairly small edits and rewrites. adding some references
moving comments to comments

The rotational evolution of cool dwarfs is mostly unconstrained after around
2-3 billion years, and the gyrochronology relations are poorly calibrated
beyond this time due to a lack of precise ages and rotation periods for old
main-sequence stars.
% However, recent evidence suggests that the relationship between rotation
% period and color/temperature/mass does not remain constant over large
% timescales, contradicting a long-held assumption that underlies many
% gyrochronology relations.
In this work we use the velocities of populations of K dwarfs to reveal the
rotational evolution of these cool stars at old ages.
We find that groups of stars, selected to be the same age using a
gyrochronology relation calibrated to the Praesepe cluster, do {\it not} have
the same velocity dispersion at all temperatures; cooler stars have greater
velocity dispersions than hotter stars.
We interpret this to mean that these cooler stars are older than the hotter
ones, which implies that the period-color relation of Praesepe is not
appropriate for old stars.
Futhermore, our results indicate that the slope of the period-color relation
actually {\it inverts}: at young ages early K stars rotate more rapidly than
late K stars, but beyond $\sim$ 2 Gyr they rotate more slowly.
These observations could be reproduced by mass-dependent dynamical heating,
where lower-mass stars experience a greater rate of dynamical heating than
higher-mass stars, however we find no evidence for mass-dependent heating
within K and M dwarfs in the \kepler\ field.

% We show however, that if mass-dependent heating is responsible, the
% differential heating rate is not constant over logarithmic time intervals
% which implies that different heating processes must be acting before and after
% ages of around 1 Gyr.
% Perhaps a more likely scenario is that the relation between rotation period
% and color changes over time, undergoing a particularly sharp transition at
% around 1 Gyr.
% If we assume that mass-dependent heating does not affect the data,

% (similar to the age of the rotation gap and the 1.1 Gyr NGC 6811
% cluster, shown to have a different period-color relation to Praesepe).
% We find that the ages of field stars with measured rotation periods do not
% exceed 4-6 Gyrs, resolving a long-standing degeneracy between age and magnetic
% braking efficiency.
% We also show that stars rotating just below the rotation period gap are
% dynamically young, ruling out the possibility that the rotation period gap is
% caused by incorrect period measurements or binary companions.

>>>>

The rotational evolution of cool dwarfs is mostly unconstrained after around
2-3 billion years, and the gyrochronology relations are poorly calibrated
beyond this time due to a lack of precise ages and rotation periods for old
main-sequence stars.
However, recent evidence suggests that the relationship between rotation
period and color/temperature/mass does not remain constant over large
timescales, contradicting a long-held assumption that underlies many
gyrochronology relations.
We use the velocities of populations of cool dwarfs to reveal the rotational
evolution of these stars at old ages.
We find that groups of stars, selected to be the same age using a
gyrochronology relation calibrated to the Praesepe cluster, do {\it not} have
the same velocity dispersion at all temperatures; cooler stars have greater
velocity dispersions than hotter stars.
We interpret this to mean that these cooler stars are older than the hotter
ones, which implies that the period-color relation of Praesepe is not
appropriate for old stars.
Futhermore, our results indicate that the slope of the period-color relation
actually {\it inverts}: at young ages early K stars rotate more rapidly than
late K stars, but beyond $\sim$ 2 Gyr they rotate more slowly.
These observations could be reproduced by mass-dependent dynamical heating,
where lower-mass stars experience a greater rate of dynamical heating than
higher-mass stars, however we find no evidence for mass-dependent heating
within K and M dwarfs in the \kepler\ field.

### 2019-08-30 paper/intro.tex
moving comments to comments

\subsection{Gyrochronology}
It is well established that the rotation periods of FGKM dwarfs increase over
time \citep{skumanich1972}.
This characteristic of main-sequence (MS) stars allows them to be dated using
their rotation periods, via gyrochronology \citep[\eg][]{kawaler1989,
pinsonneault1989, barnes2003, barnes2007, barnes2010, meibom2011, meibom2015},
which is particularly useful since the ages of MS stars are difficult to
measure via isochrone placement.
It is also well established that stars of the same age but different masses
have different rotation periods \citep[\eg][]{kraft1967, matt2012}, thought to
be caused by the deeper convective zones, and therefore stronger magnetic
dynamos (and more efficient magnetic braking) in lower-mass stars.
However, an underlying assumption behind many empirical gyrochronology
relations is that the relationships between rotation period and photometric
color\footnote{As the directly observable quantity, color is often used as a
mass proxy, and empirical gyrochronology relations are usually calibrated in
color, rather than mass or effective temperature.},
and rotation period and age are {\it separable}, meaning that the
period-color relation is the same at all ages and the period-age relation is
the same at all colors \citep[\eg][]{barnes2003, barnes2007, mamajek2008,
meibom2011, angus2015, angus2019}.
Not all gyrochronology models follow this form, particularly those motivated
based on stellar structure and evolution theory or with a Rossby number,
rather than rotation period dependence \citep[\eg][]{barnes2010, matt2012,
vansaders2016]}.
It was recently shown that old field stars rotate more rapidly than a simple,
separable gyrochronology relation would predict \citep{angus2015,
vansaders2016, vansaders2018, metcalfe2019}, and that a mass-dependent
modification to the classical \prot\ $\propto t^{\frac{1}{2}}$ spin-down law
\citep{skumanich1972} is required to reproduce the data \citep{vansaders2016,
vansaders2018}.
An even more recent analysis of the 1.1 Gyr open cluster, NGC 6811, provides
further indication that the exponent of the period-age relation is
mass and time-dependent \citep{curtis2019}.
This cluster has a flattened relationship between rotation period and color
and the G dwarfs rotate at the same rate as the K dwarfs.
% Since a separable period-age and period-color relation provided an adequate
% fit for the young stars, it was naturally assumed to apply to old stars too.

In this paper, we tested the \citet{angus2019} gyrochronology relation, a
separable relation, calibrated using the period-color relation of Praesepe (in
\gaia\ \gcolor\ color) and the period-age relation of Praesepe and the Sun.
The large number of Praesepe members with precise rotation periods from the
\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
through early M, makes it a good cluster for calibrating the period-color
relation of stars at 650 Myrs.
% This relation was calibrated by fitting a 5th-order polynomial to the relation
% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
% and the Sun.
This relation accurately describes the rotation periods of F and G stars up to
around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster with available
rotation periods), but over-predicts the rotation periods of K dwarfs in the
1.1 Gyr NGC 6811 cluster.
Very few reliable age estimates exist for K dwarfs with rotation periods older
than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
evolution of middle-aged K dwarfs is difficult to characterize\footnote{The
oldest M dwarfs with reliable ages and rotation periods are members of
Praesepe ($\sim$ 650 Myrs), so even less is known about their rotational
evolution.}.

The gyrochronal ages of the stars in our sample were calculated with the
\citep{angus2019} gyrochronology model, using de-reddened \Gaia\ \gcolor\
color and rotation periods reported by \mct.
These rotational ages are shown on the CMD in figure \ref{fig:age_gradient}.
The stars with old rotational ages, plotted in yellow hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of stars with old rotational ages are bluer
than \gcolor\ $\sim$ 1.5 dex.
This provides a first hint that the \citet{angus2019} gyrochronology relation
under-predicts the ages of low-mass stars.
There is no reason to expect the oldest stars in this sample to be the bluer
ones: M dwarfs are, on average, older than K dwarfs and are expected to remain
active for longer, so should therefore have measurable rotation periods at
older ages.
Since the \citet{angus2019} gyrochronology model, which is based on the
period-color relation of Praesepe, does not predict old ages for any M dwarfs,
it is probably under-predicting the ages of some of these low-mass stars.
It is therefore likely that old M dwarfs rotate more rapidly than this
Praesepe-based model predicts.
In what follows we use a population-based stellar age indicator, velocity
dispersion, to investigate the period-color relations at old ages in the
field.

% In this paper, we shed light on the rotational evolution of old K dwarfs using
% population-based dynamical ages.
% Unfortunately, open clusters with rotation period measurements are mostly
% young -- currently the oldest is 2.5 Gyr (NGC 6819).
% Without rotation periods for precisely dated old stars, it is extremely
% difficult to calibrate the relationship between rotation period and color at
% old ages.
% For this reason, we used a population-based stellar age indicator, velocity
% dispersion, to investigate the period-color relations at old ages in the
% field.

@@

Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
the center of the galaxy with a low out-of-plane, or vertical, velocity ($W$,
or $v_z$), just like the star-forming molecular gas observed in the disk today
\citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
% because the MW's gas disk is thin and its gas moves in circular
% orbits.
Young stars have relatively small vertical velocities, but gain momentum in
the vertical direction over time \citep[\eg][]{nordstrom2004, holmberg2007,
holmberg2009, aumer2009, casagrande2011}.
Although the cause of orbital heating is not well understood, interactions
with giant molecular clouds, spiral arms and the galactic bar are thought to
play an important role \citep[see][for a review of secular evolution in the
MW]{sellwood2014}.
Although the velocity of any individual star will only provide a weak age
constraint, the velocity dispersion of a group of stars can indicate whether,
on average, that group is old or young relative to other groups.
In this work we compare the velocity dispersions of groups of stars to
ascertain which groups are older and which younger and draw conclusions based
on the implied relative ages of populations.
% Stellar velocities have a history of being used as an age proxy, with several
% notable examples within stellar astronomy \citep[\eg][]{faherty2009,
% west2011}.
The age-velocity dispersion relations (AVRs) are still actively being
calibrated, so it is difficult to directly compare gyrochronal ages with
kinematic ones.
However, regardless of the exact relation between velocity dispersion and
stellar age, it is expected to be a monotonic relationship, therefore velocity
dispersion can be used effectively to {\it rank} groups of stars by age.

% Vertical {\it actions} are better age indicators than velocities, because
% actions are calculated by integrating angular momentum over the Milky Way's
% potential, and are therefore position invariant -- \eg\ a star will have the
% same action at periapsis and apoapsis.
% In contrast, orbital velocities are different at periapsis and apoapsis -- so
% in this sense, {\it actions} are the natural quantities to use as age proxies.
Vertical action is a better age indicator than vertical velocity
\citep{ting2019}, although still only a weak age indicator for individual
stars \citep{beane2018}, however both vertical action ($J_z$) and vertical
velocity (\vz/W) can only be calculated with full 6-dimensional position and
velocity information.
Unfortunately, most stars with measured rotation periods do not have radial
velocity (RV) measurements because they are relatively faint \kepler\ targets
($\sim$11th-18th magnitudes).
For this reason, we used velocity in the direction of galactic latitude, \vb.
The \kepler\ field is positioned at low galactic latitude (b=5-20\degrees), so
\vb\ is a close (although imperfect -- see section \ref{sec:results})
approximation to \vz.

% \subsection{The degeneracy between gyrochronology and mass-dependent heating}

% Although only calibrated using Praesepe ($\sim$ 650 Myr) and the Sun (4.56
% Gyr), the \citet{angus2019} gyrochronology relation predicts accurate ages for
% members of NGC 6819, a 2.5 Gyr open cluster.
% However there are no rotation periods for K or M dwarfs in this cluster -- its
% coolest members with rotation periods are G dwarfs.
% The period-color relation of Praesepe is nearly identical to the period-color
% relation of the Hyades, a cluster of around the same age: $\sim$ 650 years
% \citep{douglas2016, douglas2017, rebull2017}.
% However, Praesepe and the Hyades do {\it not} have the same period-color
% relation as NGC 6811, a 1.1 Gyr cluster \citep{curtis2019}.
% The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
% appears as though the K stars have `stalled' -- their spin-down has been
% halted.

% The \citet{angus2019} gyrochronology relation assumes that the rotation
% period-color relation of Praesepe is applicable to stars of all ages: the same
% polynomial relation fit to Praesepe is used to describe the period-color
% relation for all stars.
% However, the NGC 6811 cluster suggests that this assumption does not hold for
% K dwarfs past around 1 Gyr.
% If the period-color relation of the \citet{angus2019} gyrochronology model
% {\it were} a perfect model for the rotational evolution of stars, then groups
% of stars selected to be similar ages using this relation should fall on an
% isochrone on a CMD.
% % and have the same velocity dispersion across all colors.
% Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
% variations in metallicity and extinction, blur out the main sequence enough
% that differences between stars in different age and rotation bins are not
% easily discernable on the CMD, although there is still a general age and
% rotation period gradient on the CMD, as seen in figures \ref{fig:CMD_cuts} and
% \ref{fig:age_gradient}.
% For this reason, we chose to use {\it kinematic} `isochrones', instead of
% magnitude and color-based isochrones: although kinematics as an age indicator
% is not necessarily as well calibrated to an absolute age scale as CMD
% position, it can be very sensitive to {\it differences} in the ages of stellar
% populations.

% Isochrones and stellar evolution tracks are highly dependent on choices made
% about input physics and assumptions.
% % and have often been calibrated using
% % different types of stars.
% As a result, different sets of models can have very different shapes on the
% CMD, particularly at low masses.
% In fact, only empirical models, not physical ones, are currently able to
% reproduce the CMD positions of M dwarfs.
% Instead of relying on CMD position to age-date groups of stars, we opted to
% explore age trends via kinematics.
% Kinematic age-dating has the advantage of being relatively model independent,
% or at least, having a very simple model: that velocity dispersion increases
% over time.
% This means that it is relatively easy to rank groups of stars by age: older
% groups have a larger velocity dispersion.

This paper is laid out as follows: in section \ref{sec:the_data} we describe
our sample selection process and the methods used to calculated stellar
velocities.
In section \ref{sec:results} we use stellar kinematics to investigate the
relationship between rotation period, age and color/\teff\ in field stars.
In section \ref{sec:mass-dependent-heating} we show that neither
mass-dependent heating nor the selection function is likely to have a strong
affect on our sample.
Finally, in section \ref{sec:period_gap} we discuss the implications of our
results.

>>>>

\subsection{Gyrochronology}
It is well established that the rotation periods of FGKM dwarfs increase over
time \citep{skumanich1972}.
This characteristic of main-sequence (MS) stars allows them to be dated using
their rotation periods, via gyrochronology \citep[\eg][]{kawaler1989,
pinsonneault1989, barnes2003, barnes2007, barnes2010, meibom2011, meibom2015},
which is particularly useful since the ages of MS stars are difficult to
measure via isochrone placement.
It is also well established that stars of the same age but different masses
have different rotation periods \citep[\eg][]{kraft1967, matt2012}, thought to
be caused by the deeper convective zones, and therefore stronger magnetic
dynamos (and more efficient magnetic braking) in lower-mass stars.
However, an underlying assumption behind many empirical gyrochronology
relations is that the relationships between rotation period and photometric
color\footnote{As the directly observable quantity, color is often used as a
mass proxy, and empirical gyrochronology relations are usually calibrated in
color, rather than mass or effective temperature.},
and rotation period and age are {\it separable}, meaning that the
period-color relation is the same at all ages and the period-age relation is
the same at all colors \citep[\eg][]{barnes2003, barnes2007, mamajek2008,
meibom2011, angus2015, angus2019}.
Not all gyrochronology models follow this form, particularly those motivated
based on stellar structure and evolution theory or with a Rossby number,
rather than rotation period dependence \citep[\eg][]{barnes2010, matt2012,
vansaders2016]}.
It was recently shown that old field stars rotate more rapidly than a simple,
separable gyrochronology relation would predict \citep{angus2015,
vansaders2016, vansaders2018, metcalfe2019}, and that a mass-dependent
modification to the classical \prot\ $\propto t^{\frac{1}{2}}$ spin-down law
\citep{skumanich1972} is required to reproduce the data \citep{vansaders2016,
vansaders2018}.
An even more recent analysis of the 1.1 Gyr open cluster, NGC 6811, provides
further indication that the exponent of the period-age relation is
mass and time-dependent \citep{curtis2019}.
This cluster has a flattened relationship between rotation period and color
and the G dwarfs rotate at the same rate as the K dwarfs.

In this paper, we tested the \citet{angus2019} gyrochronology relation, a
separable relation, calibrated using the period-color relation of Praesepe (in
\gaia\ \gcolor\ color) and the period-age relation of Praesepe and the Sun.
The large number of Praesepe members with precise rotation periods from the
\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
through early M, makes it a good cluster for calibrating the period-color
relation of stars at 650 Myrs.
This relation accurately describes the rotation periods of F and G stars up to
around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster with available
rotation periods), but over-predicts the rotation periods of K dwarfs in the
1.1 Gyr NGC 6811 cluster.
Very few reliable age estimates exist for K dwarfs with rotation periods older
than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
evolution of middle-aged K dwarfs is difficult to characterize\footnote{The
oldest M dwarfs with reliable ages and rotation periods are members of
Praesepe ($\sim$ 650 Myrs), so even less is known about their rotational
evolution.}.

The gyrochronal ages of the stars in our sample were calculated with the
\citep{angus2019} gyrochronology model, using de-reddened \Gaia\ \gcolor\
color and rotation periods reported by \mct.
These rotational ages are shown on the CMD in figure \ref{fig:age_gradient}.
The stars with old rotational ages, plotted in yellow hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of stars with old rotational ages are bluer
than \gcolor\ $\sim$ 1.5 dex.
This provides a first hint that the \citet{angus2019} gyrochronology relation
under-predicts the ages of low-mass stars.
There is no reason to expect the oldest stars in this sample to be the bluer
ones: M dwarfs are, on average, older than K dwarfs and are expected to remain
active for longer, so should therefore have measurable rotation periods at
older ages.
Since the \citet{angus2019} gyrochronology model, which is based on the
period-color relation of Praesepe, does not predict old ages for any M dwarfs,
it is probably under-predicting the ages of some of these low-mass stars.
It is therefore likely that old M dwarfs rotate more rapidly than this
Praesepe-based model predicts.
In what follows we use a population-based stellar age indicator, velocity
dispersion, to further investigate the gyrochronology relations at old ages in
the field.

@@

Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
the center of the galaxy with a low out-of-plane, or vertical, velocity ($W$,
or $v_z$), just like the star-forming molecular gas observed in the disk today
\citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
Young stars have relatively small vertical velocities, but gain momentum in
the vertical direction over time \citep[\eg][]{nordstrom2004, holmberg2007,
holmberg2009, aumer2009, casagrande2011}.
Although the cause of orbital heating is not well understood, interactions
with giant molecular clouds, spiral arms and the galactic bar are thought to
play an important role \citep[see][for a review of secular evolution in the
MW]{sellwood2014}.
Although the velocity of any individual star will only provide a weak age
constraint, the velocity dispersion of a group of stars can indicate whether,
on average, that group is old or young relative to other groups.
In this work we compare the velocity dispersions of groups of stars to
ascertain which groups are older and which younger and draw conclusions based
on the implied relative ages of populations.
The age-velocity dispersion relations (AVRs) are still actively being
calibrated, so it is difficult to directly compare gyrochronal ages with
kinematic ones.
However, regardless of the exact relation between velocity dispersion and
stellar age, it is expected to be a monotonic relationship, therefore velocity
dispersion can be used effectively to {\it rank} groups of stars by age.

Vertical action is thought to be a better age indicator than vertical velocity
\citep{ting2019}, although still only a weak age indicator for individual
stars \citep{beane2018}, however both vertical action ($J_z$) and vertical
velocity (\vz/W) can only be calculated with full 6-dimensional position and
velocity information.
Unfortunately, most stars with measured rotation periods do not have radial
velocity (RV) measurements because they are relatively faint \kepler\ targets
($\sim$11th-18th magnitudes).
For this reason, we used velocity in the direction of galactic latitude, \vb.
The \kepler\ field is positioned at low galactic latitude (b=5-20\degrees), so
\vb\ is a close (although imperfect -- see section \ref{sec:results})
approximation to \vz.

This paper is laid out as follows: in section \ref{sec:the_data} we describe
our sample selection process and the methods used to calculated stellar
velocities.
In section \ref{sec:results} we use stellar kinematics to investigate the
relationship between rotation period, age and color/\teff\ in the field.
In section \ref{sec:mass-dependent-heating} we show that neither
mass-dependent heating nor the selection function is likely to have a strong
affect on our sample, and in section \ref{sec:period_gap} we discuss the
implications of our results.

### 2019-08-30 paper/method.tex
moving comments to comments

% \begin{figure}
%   \caption{
% A \gaia\ color magnitude diagram showing the \citet{mcquillan2014} sample with
%     extinction-corrected magnitudes, colored by rotation period.
% We excluded photometric binaries and subgiants from our analysis by removing
% stars above the two dashed lines.
% % The rotation periods of binaries and subgiants do not follow a Skumanich-like
% % braking law.
% % The rotation period gradient across the main sequence is visible by eye in
% %     this figure: young, rapidly rotating stars are located below the old,
% %     slowly rotating stars.
% % since we rely on a simple scaling
% % between rotation period and age to make the argument that the rotation period
% % is not caused by incorrect period measurements.
% }
%   \centering
%     \includegraphics[width=1\textwidth]{CMD_cuts}
% \label{fig:CMD_cuts}
% \end{figure}
% To explore the age of this stellar population from a rotation standpoint, it
% was first necessary to remove visual binaries and subgiants from the sample.
% The rotational evolution of these two types of stars is generally different to
% that of single stars which more usually follow a Skumanich-like spin-down law.
% Tidal and magnetic interactions between the two components of a binary system
% can influence the rotation periods of both stars, and the expanding envelopes
% of subgiants drive rapid spin-down through conservation of angular momentum.
Visual binaries and subgiants were removed from the sample by applying cuts to
the color-magnitude diagram (CMD), shown in figure \ref{fig:age_gradient}.
A 6th-order polynomial was fit to the main sequence and raised by 0.27 dex to
approximate the division between single stars and visual binaries.
All stars above this line were removed from the sample.
Subgiants were also removed by eliminating stars brighter than 6th magnitude
in \gaia\ G-band.
\begin{figure}
  \caption{
Dereddened MS \kepler\ stars with \mct\ rotation periods on the \gaia\ CMD.
We excluded visual binaries by removing stars above the dashed line.
    Points are colored by their gyrochronal
    age, according to the
    \citet{angus2019} gyrochronology relation.
    A general age gradient is visible across the main sequence.
}
  \centering
    \includegraphics[width=1\textwidth]{age_gradient_combo}
\label{fig:age_gradient}
\end{figure}

@@

% \begin{figure}
%   \caption{
% the rotation periods of stars in the \citet{mcquillan2014} sample vs.
% effective temperature.
% black points are dwarf stars identified as non-photometric binaries, hotter
% than 4800 k.
% blue circle points are non-photometric binary dwarfs, cooler than 4800 k, with
% a rotation period and \gaia\ color indicating they are older than 1.1 gyr.
% orange squares are stars that with rotation periods that fall just below the
% gap: they have rotation-ages between 0.5 and 1.1 gyrs.
% green triangles are stars with rotation periods faster than the main envelop
% of stars.
% these are probably binaries whose rotation periods are synchronized to their
% orbits and have been spun-up via tidal interactions.
% }
%   \centering
%     \includegraphics[width=1\textwidth]{period_teff}
% \label{fig:period_teff}
% \end{figure}

>>>>

Visual binaries and subgiants were removed from the sample by applying cuts to
the color-magnitude diagram (CMD), shown in figure \ref{fig:age_gradient}.
A 6th-order polynomial was fit to the main sequence and raised by 0.27 dex to
approximate the division between single stars and visual binaries.
All stars above this line were removed from the sample.
Subgiants were also removed by eliminating stars brighter than 6th magnitude
in \gaia\ G-band.
\begin{figure}
  \caption{
Dereddened MS \kepler\ stars with \mct\ rotation periods on the \gaia\ CMD.
We excluded visual binaries by removing stars above the dashed line.
    Points are colored by their gyrochronal
    age, according to the
    \citet{angus2019} gyrochronology relation.
    A general age gradient is visible across the main sequence.
}
  \centering
    \includegraphics[width=1\textwidth]{age_gradient_combo}
\label{fig:age_gradient}
\end{figure}

@@

\subsection{Establishing \vb\ as an age proxy}
\label{sec:mass-dependent-heating}

If lower-mass stars experience greater velocity changes when gravitationally
perturbed and are dynamically heated more efficiently than higher-mass stars,
that means velocity dispersion is a function of {\it both} age and mass and
cannot be straightforwardly interpreted as an age proxy.
Mass-dependent dynamical heating has not been unambiguously observed in the
galactic disk because of the strong anti-correlation between stellar mass and
stellar age.
Less massive stars do indeed have larger velocity dispersions, however they
are also older on average.
This mass-age degeneracy is highly reduced in M dwarfs because their
main-sequence lifetimes are longer than the age of the Universe, however no
evidence for mass-dependent heating has been found in M dwarfs
\citep{faherty2009}.

To investigate whether mass-dependent heating could be acting on the \kepler\
sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
whose MS lifetimes exceed around 11 Gyrs and are therefore representative of
the initial mass function.
We could not perform this analysis on the \mct\ sample, because only stars
with {\it detectable} rotation periods appear there.
Since lower-mass stars stay active for longer it is likely that there are more
old stars (with larger velocity dispersions) at low masses in this sample.
We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
greater than 1.2 (corresponding to an effective temperature $\lesssim$
4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
We also eliminated visual binaries by removing stars above a 6th order
polynomial, fit to the MS on the \gaia\ CMD.
We then applied the quality cuts described in section \ref{sec:the_data}.
To search for evidence of mass-dependent heating we calculated the (\vb)
velocity dispersion of stars in effective temperature bins.
Again, sigma clipping was used to remove high velocity outliers before
calculating the standard deviation of stars in each bin.

Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion
as a function of effective temperature (calculated by transforming dereddened
\gaia\ colors using equation \ref{eq:curtis}).
Velocity dispersion very slightly {\it decreases} with decreasing temperature,
(the opposite of the trend expected for mass-dependent heating) however the
slope is only inconsistent with zero at the 1.3 $\sigma$ level.
This trend may be due to a selection bias: cooler stars are fainter and
therefore typically closer, with smaller heights above the galactic plane and
smaller velocities.
The essential point here is that we do not find  evidence for mass-dependent
heating acting on stars in the \kepler\ field, meaning \sigmavb\ {\it can} be
used as an age proxy.
We performed the same analysis on the 537 stars in this sample with RVs using
{\it vertical} velocity (\vz) and found the slope of the velocity
dispersion-temperature relation was consistent with zero.

Since we used \vb, not \vz\ in our analysis, it is possible that an
anti-correlation between \teff\ and galactic latitude, $b$, could influence
our results: stars at higher latitudes have additional velocity components in
the ${\bf x}$ and ${\bf y}$ directions, which could increase \vb\ but not \vz.
Again however, since the relationship between \sigmavb\ and \teff\ is
positively, not negatively correlated for cool stars in the \kepler\ field,
this effect is probably too small to influence our results.

\begin{figure}
  \caption{
      Top: Stellar velocity (\vb) as a function of \teff\ for
      \kepler\ K and M dwarfs.
Vertical lines indicate different \teff-groupings used to calculate velocity
    dispersion.
Pink stars were not included in velocity dispersion calculations as they were
    either removed as outliers during a sigma clipping process, or they lie at
    the sparcely populated, extremely cool end of the temperature range.
    Velocity dispersion and \teff\ are slightly positively correlated, likely
    due to a brightness-related selection bias, indicating that mass-dependent
    heating does not significantly affect low-mass stars in the \kepler\
    field.
}
  \centering
    \includegraphics[width=1\textwidth]{vb_vs_teff}
\label{fig:vb_vs_teff}
\end{figure}
Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
strongly affect the \mct\ sample of rotating \kepler\ stars.
For this reason, we assume that age difference is the major cause of velocity
dispersion differences between groups.
In other words, (\vb) velocity dispersion is a reliable age proxy for the
\mct\ sample.
